\chapter{Asymmetry Phenotypic Analysis}\label{chap:asymmetry}
\section{Statistical analysis of asymmetry components}
Bilateral asymmetry is mainly described using three components in literature \cite{klingenberg2002}\cite{Vingerhoets2021}; \acf{da}, the focus of this study, corresponds to the hemispheric side effect; antisymmetry, which is related to the effect where sidedness is random in a population (i.e. left-right randomly switches to right-left), is not observed in the human cerebral cortex, in contrast to other internal organs positions, or organisms \cite{Neubauer2020}; \acf{fa}, encompasses any random developmental and environmental effects, that cannot be explained with the existing knowledge. The observed deviations can be statistically linearly modeled as products of two effects, the hemisphere side studied and the individual specimen analyzed, as well as their interaction \cite{klingenberg2002}. Formally, based on \cite{VanDongen1999} assuming the presence of replications of the observation per individual, to account for technical error, a mixed linear model representing the aforementioned dependencies is defined as:
\begin{equation}
	Y_{ijk} = \mu + \beta + I_i + S_{ij} + E_{ijk}
\end{equation}
where $Y_{ijk}$ is the phenotype of the i-th individual, from the j-th side, under the k-th replication, $\mu$ and $\beta$ are the fixed intercept and fixed side effect respectively, $I_i\sim\mathcal{N}(0,\sigma^2_{ind})$ is the random individual effect,  $S_{ij}\sim\mathcal{N}(0,\sigma^2_{FA})$ is the random side and individual specific effect, matched to \ac{fa}, and $E_{ijk}\sim\mathcal{N}(0,\sigma^2_{ME})$ is the measurement error. Replications are necessary in such a study, in order to differentiate \ac{fa} effect from the measurement error. Given this definition, a way to measure the statistical significance is through an F-test applied on 2-way \ac{anova}, to relate the \ac{rss} ratios of effects to observable error terms, and of fluctuating effect to the measurement error. Extra care needs to be given on the determination of the \ac{dof} of each term, given the preprocessing applied to bring the hemispheres surfaces into Kendall shape space. Those are extracted from the rigorous work in \cite{klingenberg2002}. Given that the analysis is performed on a pair of symmetric objects, and not on a single symmetric object, this configuration is named matching asymmetry analysis. In order to avoid further de facto assumptions, regarding the distribution of TO BE CONTINUED
\section{Phenotypic partitioning}\label{sec:hsc}
\Acf{hsc} is an unsupervised method of iterative partitioning, that makes use of the distance matrix eigenvectors \cite{Ng2002}. It results into a binary tree structure (i.e. each parent shape is partitioned into two children). In the current study, a level-4 partitioning is performed, resulting into 31 partitions. Subsequently, they are transformed to the corresponding principal components that explain 80\% of the variance, for reasons of further dimensionality reduction. TO BE CONTINUED